---
title: "R scripts"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: show
    highlight: tango
    df_print: paged
    css: assets/custom.css
    includes:
      after_body: assets/after_body.html
---


This page collects reusable R utilities (installation helpers, summary/stat functions, trait workflows, PCA, TPD utilities, and assorted snippets).  
All chunks are shown for copy/paste but **not evaluated** (`eval = FALSE`).

## Contents {.tabset .tabset-fade .tabset-pills}

### Library

```{r library, eval=FALSE}
packages <- c(
  "ade4","ape","berryFunctions","betapart","biscale","cowplot","PCA_TRAITS.table",
  "dplyr","funrar","geiger","ggplot2","ggpubr","lsmeans","missForest","motmot",
  "multcomp","mvMORPH","paleotree","pals","paran","phylobase","phytools","picante",
  "plotly","plotrix","psych","quanteda","ratematrix","RColorBrewer","readr","rgbif",
  "rnaturalearth","rredlist","sf","shape","stats","tidyr","TPD","vegan",
  "VennDiagram","viridis","wesanderson"
)

installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}
# Packages loading
lapply(packages, library, character.only = TRUE)
```

### Statistics

```{r statistics, eval=FALSE}
distrib <- function(x, k = 6) {
  # NA removing
  x <- x[is.na(x) == FALSE]

  Nx <- length(x)
  sum <- round(summary(x), k)
  res <- c(Nx, sum)
  names(res) <- c("N", names(sum))
  return(res)
} # end of function distrib

hms_span <- function(start, end) {
  dsec <- as.numeric(difftime(end, start, unit = "secs"))
  hours <- floor(dsec / 3600)
  minutes <- floor((dsec - 3600 * hours) / 60)
  seconds <- dsec - 3600 * hours - 60 * minutes
  paste0(
    sapply(c(hours, minutes, seconds), function(x) {
      formatC(x, width = 2, format = "d", flag = "0")
    }),
    collapse = ":"
  )
}

sesandpvalues_bis <- function(obs, rand, nreps, probs = c(0.025, 0.975), rnd = 2) {
  SES <- (obs - mean(rand)) / sd(rand)
  pValsSES <- rank(c(obs, rand))[1] / (length(rand) + 1)
  results <- round(c(obs, SES, mean(rand), quantile(rand, prob = probs), pValsSES, nreps), rnd)
  names(results) <- c("Observed", "SES", "MeanRd", "CI025Rd", "CI975Rd", "Pval", "Nreps")
  return(results)
}

extract_t_test <- function(x, y, ...) {
  z <- t.test(x, y)
  stat <- z$statistic
  pval <- z$p.value
  pvalStars <- ifelse(pval > 0.05, ".",
                      ifelse(pval > 0.01, "*",
                             ifelse(pval > 0.001, "**", "***")))
  est <- z$estimate
  sdv <- c(sd(x), sd(y))

  res <- data.frame(est[1], sdv[1], est[2], sdv[2], stat, pval, pvalStars)
  res[1:4] <- round(as.numeric(res[1:4]), 2)
  names(res) <- c("Mean x", "sd x", "Mean y", "sd y", "t", "Pvalues", "P")
  return(res)
}

extract_indices_list <- function(ind) {
  indSES <- matrix(
    NA, nr = ncol(ind), nc = 7,
    dimnames = list(
      colnames(ind),
      c("Observed", "SES", "MeanRd", "CI025Rd", "CI975Rd", "Pval", "Nreps")
    )
  )
  for (i in 1:nrow(indSES)) {
    indSES[i, ] <- sesandpvalues_bis(
      obs = ind[1, i],
      rand = ind[-1, i],
      nreps = length(ind[, 1]) - 1,
      probs = c(0.025, 0.975),
      rnd = 2
    )
  }
  rownames(indSES) <- gsub("\\.", " ", rownames(indSES))
  return(indSES)
}

extract_indices_change <- function(ind) {
  indSES <- matrix(
    NA, nr = ncol(ind), nc = 7,
    dimnames = list(
      colnames(ind),
      c("Observed", "SES", "MeanRd", "CI025Rd", "CI975Rd", "Pval", "Nreps")
    )
  )
  for (i in 1:nrow(indSES)) {
    indSES[i, ] <- sesandpvalues_bis(
      obs = ((ind[2, i] - ind[1, i]) / ind[1, i]) * 100,
      rand = ((ind[-c(1:2), i] - ind[1, i]) / ind[1, i]) * 100,
      nreps = length(ind[, 1]) - 2,
      probs = c(0.025, 0.975),
      rnd = 2
    )
  }
  rownames(indSES) <- gsub("\\.", " ", rownames(indSES))
  return(indSES)
}

extract_indices_change_log <- function(ind) {
  indSES <- matrix(
    NA, nr = ncol(ind), nc = 7,
    dimnames = list(
      colnames(ind),
      c("Observed", "SES", "MeanRd", "CI025Rd", "CI975Rd", "Pval", "Nreps")
    )
  )
  for (i in 1:nrow(indSES)) {
    indSES[i, ] <- sesandpvalues_bis(
      obs = log(ind[2, i] / ind[1, i]),
      rand = log(ind[-c(1:2), i] / ind[1, i]),
      nreps = length(ind[, 1]) - 2,
      probs = c(0.025, 0.975),
      rnd = 2
    )
  }
  rownames(indSES) <- gsub("\\.", " ", rownames(indSES))
  return(indSES)
}
```

### Sorting data

```{r sorting_data, eval=FALSE}
standard_names <- function(Traits, nrowTraits, source = 163, dim = NA) {
  require(taxize)
  standNamesTraits <- as.data.frame(
    matrix(
      NA,
      nrow = nrowTraits,
      ncol = 1,
      dimnames = list(rownames(Traits), "IUCNName")
    )
  )
  batchSize <- 1000

  for (i in 1:(ceiling(nrowTraits / batchSize))) {
    cat(paste0("\nBatch ", i, " out of ", (ceiling(nrowTraits / batchSize)), "\n"))
    rowsSelect <- ((i - 1) * batchSize + 1):min((i * batchSize), nrowTraits)
    taxoAux <- as.data.frame(
      taxize::gnr_resolve(
        names = rownames(standNamesTraits)[rowsSelect],
        preferred_data_sources = source, # 163 is IUCN
        best_match_only = TRUE,
        canonical = TRUE
      )
    )
    standNamesTraits[taxoAux$user_supplied_name, "IUCNName"] <- taxoAux$matched_name2
  }

  standNamesTraits[, "IUCNName"] <- gsub(
    x = standNamesTraits[, "IUCNName"],
    pattern = " ",
    replacement = "_"
  )
  naIUCN <- which(is.na(standNamesTraits[, "IUCNName"]))
  standNamesTraits[naIUCN, "IUCNName"] <- rownames(standNamesTraits)[naIUCN]

  # Some species could not be resolved into IUCN naming. Remove them:
  yesName <- which(!is.na(standNamesTraits[, "IUCNName"]))
  if (dim == 1) {
    return(Traits[yesName])
  } else {
    return(Traits[yesName, ])
  }
}

addToPhylo <- function(phylogenyTraits, GROUPTraits) {
  phylogenyTraits <- force.ultrametric(tree = phylogenyTraits, method = c("extend"))
  tipsToAdd <- which(!rownames(GROUPTraits) %in% phylogenyTraits$tip.label)
  namesToAdd <- rownames(GROUPTraits)[tipsToAdd]
  genusToAdd <- unlist(strsplit(namesToAdd, "_"))[
    which(1:length(strsplit(namesToAdd, "_")) %% 2 == 1)
  ]
  phylogenyTraitsAdd <- phylogenyTraits

  for (add in 1:length(namesToAdd)) {
    cat(paste("\rADDING SP TO GENUS IN PHYLOGENY", add, "/", length(namesToAdd), "\r"))
    phylogenyTraitsAdd <- add.species.to.genus(
      tree = phylogenyTraitsAdd,
      species = namesToAdd[add],
      genus = genusToAdd[add],
      where = "root"
    )
  }
  return(phylogenyTraitsAdd)
}

RedToTraits <- function(IUCNGROUPtoTraits) {
  RedToTraits <- matrix(
    rep(NA, nrow(IUCNGROUPtoTraits)),
    ncol = 1,
    dimnames = list(IUCNGROUPtoTraits$Species, "IUCN")
  )

  for (i in 1:nrow(IUCNGROUPtoTraits)) {
    selRows <- which(IUCNGROUPtoTraits$Species == IUCNGROUPtoTraits$Species[i])
    if (length(selRows) > 1) {
      dataAux <- IUCNGROUPtoTraits[selRows, ]
      if (any(is.na(dataAux$infra_name))) {
        coger <- which(is.na(dataAux$infra_name))
        if (length(coger) == 1) {
          RedToTraits[IUCNGROUPtoTraits$Species[i], "IUCN"] <- as.character(dataAux$categorySimp[coger])
        } else {
          if (length(unique(dataAux$categorySimp[coger])) == 1) {
            RedToTraits[IUCNGROUPtoTraits$Species[i], "IUCN"] <-
              as.character(unique(dataAux$categorySimp[coger]))
          } else {
            RedToTraits[IUCNGROUPtoTraits$Species[i], "IUCN"] <-
              as.character(sort(unique(dataAux$categorySimp))[1])
          }
        }
      } else {
        RedToTraits[IUCNGROUPtoTraits$Species[i], "IUCN"] <-
          as.character(sort(unique(dataAux$categorySimp))[1])
      }
    } else {
      RedToTraits[IUCNGROUPtoTraits$Species[i], "IUCN"] <-
        as.character(IUCNGROUPtoTraits[selRows, "categorySimp"])
    }
  }
  return(RedToTraits)
}

imputation_trait <- function(GROUPData, selectedTraits, shortnames = c("am", "bs", "ls", "os")) {
  require(missForest)
  columnsTraits <- 1:(which(colnames(GROUPData) == "Eigen.1") - 1)
  GROUPData[, columnsTraits] <- log10(GROUPData[, columnsTraits])

  # NOTE: All traits are kept for imputation
  columsImputation <- 1:(which(colnames(GROUPData) == "IUCN") - 1)
  GROUPTraitslogImputed <- as.matrix(missForest(xmis = GROUPData[, columsImputation])$ximp)

  GROUPTraitsImputed <- GROUPTraitsMissing <- GROUPData
  GROUPTraitsImputed[, columnsTraits] <- GROUPTraitslogImputed[, columnsTraits]

  GROUPTraitsMissing <- GROUPTraitsMissing[, selectedTraits]
  GROUPTraitsImputed <- GROUPTraitsImputed[, selectedTraits]
  colnames(GROUPTraitsMissing) <- colnames(GROUPTraitsImputed) <- shortnames

  GROUPOtherInfo <- GROUPData[, (max(columnsTraits) + 1):ncol(GROUPData)]
  GROUPTraitsImputed <- data.frame(GROUPTraitsImputed, IUCN = GROUPData$IUCN)
  GROUPTraitsMissing <- data.frame(GROUPTraitsMissing, IUCN = GROUPData$IUCN)

  return(list(GROUPTraitsImputed, GROUPTraitsMissing))
}
```

### PCA

```{r pca, eval=FALSE}
# Perform PCA
# TRAITS: table of traits sp x tr : MUST BE SCALED AND log10
make.PCA <- function(TRAITS, dimensionsAux, savePCA) {
  paranAux <- paran(TRAITS)
  PCA_TRAITS <- list()
  PCA_TRAITS$traits <- TRAITS
  PCA_TRAITS$PCA <- princomp(PCA_TRAITS$traits)
  PCA_TRAITS$dimensions <- dimensionsAux
  PCA_TRAITS$variance <- (summary(PCA_TRAITS$PCA)[1][[1]]^2)[1:dimensionsAux] /
    sum(summary(PCA_TRAITS$PCA)[1][[1]]^2)
  PCA_TRAITS$loadings <- PCA_TRAITS$PCA$loadings
  PCA_TRAITS$traitsUse <- data.frame(PCA_TRAITS$PCA$scores[, 1:PCA_TRAITS$dimensions])
  if (!is.na(savePCA)) {
    saveRDS(PCA_TRAITS, savePCA)
  }
  return(PCA_TRAITS)
}

# Possible to turn it as function for all groups (Nat Comm 2021)
make.PCA.groups <- function(traitsImputed, traitsMissing,
                            group = c("Plants","Birds","Mammals","Reptiles","Amphibians","Fishes")) {
  require(paran)
  paranAux <- paran(traitsImputed[, 1:(ncol(traitsImputed) - 1)])
  if (paranAux$Retained < 2) paranAux$Retained <- 2
  dimensionsAux <- paranAux$Retained

  PCAImpute <- list()
  PCAImpute$traits <- traitsImputed[, 1:(ncol(traitsImputed) - 1)]
  PCAImpute$traitsMissing <- traitsMissing[, 1:(ncol(traitsImputed) - 1)]
  PCAImpute$nImputed <- rowSums(is.na(PCAImpute$traitsMissing))
  PCAImpute$PCA <- princomp(traitsImputed[, 1:(ncol(traitsImputed) - 1)])
  PCAImpute$dimensions <- dimensionsAux
  PCAImpute$variance <- (summary(PCAImpute$PCA)[1][[1]]^2)[1:dimensionsAux] /
    sum(summary(PCAImpute$PCA)[1][[1]]^2)

  if (group %in% c("Birds","Mammals")) { # Mammals & Birds
    multPC1 <- ifelse(PCAImpute$PCA$loadings["bm", "Comp.1"] > 0, 1, -1)
    PCAImpute$PCA$scores[, 1] <- multPC1 * PCAImpute$PCA$scores[, 1]
    PCAImpute$PCA$loadings[, "Comp.1"] <- multPC1 * PCAImpute$PCA$loadings[, "Comp.1"]
  }

  if (group == "Reptiles") { # Reptiles
    multPC2 <- ifelse(PCAImpute$PCA$loadings["inc", "Comp.2"] > 0, -1, 1)
    PCAImpute$PCA$scores[, 2] <- multPC2 * PCAImpute$PCA$scores[, 2]
    PCAImpute$PCA$loadings[, "Comp.2"] <- multPC2 * PCAImpute$PCA$loadings[, "Comp.2"]
  }

  if (group == "Fishes") { # Fishes
    multPC1 <- ifelse(PCAImpute$PCA$loadings["svl", "Comp.1"] > 0, 1, -1)
    multPC2 <- ifelse(PCAImpute$PCA$loadings["svl", "Comp.2"] < 0, 1, -1)
    PCAImpute$PCA$scores[, 1] <- multPC1 * PCAImpute$PCA$scores[, 1]
    PCAImpute$PCA$loadings[, "Comp.1"] <- multPC1 * PCAImpute$PCA$loadings[, "Comp.1"]
    PCAImpute$PCA$scores[, 2] <- multPC2 * PCAImpute$PCA$scores[, 2]
    PCAImpute$PCA$loadings[, "Comp.2"] <- multPC2 * PCAImpute$PCA$loadings[, "Comp.2"]
  }

  PCAImpute$loadings <- PCAImpute$PCA$loadings
  PCAImpute$traitsUse <- data.frame(PCAImpute$PCA$scores[, 1:PCAImpute$dimensions])

  treathCat <- c("EX", "EW", "CR", "EN", "VU")
  noThreatCat <- c("LC", "NT")
  PCAImpute$Threat <- with(traitsImputed,
                           ifelse(IUCN %in% treathCat, 1,
                                  ifelse(IUCN %in% noThreatCat, 0, NA)))
  return(PCAImpute)
}
```

### Functional diversity

#### TPDsMean without ITV

```{r tpd_mean_without_itv, eval=FALSE}
# PCA_TRAITS: object from PCA previously done
dimensions <- PCA_TRAITS$dimensions
traitsUSE <- PCA_TRAITS$traitsUse
colnames(traitsUSE) <- paste0("Comp.", 1:dimensions)
gridSize <- ifelse(dimensions == 4, 30, 100)
sdTraits <- sqrt(diag(Hpi.diag(traitsUSE)))

TPDsAux <- TPDsMean(
  species = rownames(traitsUSE),
  means = traitsUSE,
  sds = matrix(rep(sdTraits, nrow(traitsUSE)), byrow = TRUE, ncol = dimensions),
  alpha = 0.95,
  n_divisions = gridSize
)

make.TPD.2D.high.def <- function(targetGroupName, PCAImpute, alphaUse = 0.95,
                                 saveFile = paste0(getwd(), "/TPDs2DHighDef.rds")) {
  require(TPD)
  cat(paste("\n ESTIMATING TPDs for: ", targetGroupName, "\n"))
  dimensions <- 2
  traitsUSE <- PCAImpute$traitsUse[, 1:dimensions]
  colnames(traitsUSE) <- paste0("Comp.", 1:2)
  gridSize <- 200
  sdTraits <- sqrt(diag(Hpi.diag(traitsUSE)))

  TPDsAux <- TPDsMean(
    species = rownames(traitsUSE),
    means = traitsUSE,
    sds = matrix(rep(sdTraits, nrow(traitsUSE)), byrow = TRUE, ncol = dimensions),
    alpha = alphaUse,
    n_divisions = gridSize
  )
  saveRDS(TPDsAux, saveFile)
  TPDsAux <- NULL
  gc()
}

make.TPD.2D.low.def <- function(targetGroupName, PCAImpute, alphaUse = 0.95,
                                saveFile = paste0(getwd(), "/TPDs2DLowDef.rds")) {
  require(TPD)
  cat(paste("\n ESTIMATING TPDs for: ", targetGroupName, "\n"))
  dimensions <- 2
  traitsUSE <- PCAImpute$traitsUse[, 1:dimensions]
  colnames(traitsUSE) <- paste0("Comp.", 1:2)
  gridSize <- 100
  sdTraits <- sqrt(diag(Hpi.diag(traitsUSE)))

  TPDsAux <- TPDsMean(
    species = rownames(traitsUSE),
    means = traitsUSE,
    sds = matrix(rep(sdTraits, nrow(traitsUSE)), byrow = TRUE, ncol = dimensions),
    alpha = alphaUse,
    n_divisions = gridSize
  )
  saveRDS(TPDsAux, saveFile)
  TPDsAux <- NULL
  gc()
}

make.TPDLarge <- function(targetGroupName, PCAImpute, alphaUse = 0.95, TPDsMean_large,
                          saveFile = paste0(getwd(), "/TPDsLarge.rds")) {
  cat(paste("\n ESTIMATING TPDs for: ", targetGroupName, "\n"))
  dimensions <- PCAImpute$dimensions
  traitsUSE <- PCAImpute$traitsUse
  colnames(traitsUSE) <- paste0("Comp.", 1:dimensions)
  gridSize <- ifelse(dimensions == 4, 30, 100)
  sdTraits <- sqrt(diag(Hpi.diag(traitsUSE)))

  TPDsAux <- TPDsMean_large(
    species = rownames(traitsUSE),
    means = traitsUSE,
    sds = matrix(rep(sdTraits, nrow(traitsUSE)), byrow = TRUE, ncol = dimensions),
    alpha = alphaUse,
    n_divisions = gridSize
  )
  saveRDS(TPDsAux, saveFile)
  TPDsAux <- NULL
  gc()
}
```

#### Function TPD package updated

```{r tpd_updated_functions, eval=FALSE}
TPDRichness <- function(TPDc = NULL, TPDs = NULL) {
  if (is.null(TPDc) & is.null(TPDs)) {
    stop("At least one of 'TPDc' or 'TPDs' must be supplied")
  }
  if (!is.null(TPDc) & class(TPDc) != "TPDcomm") {
    stop("Class mismatch: please specify if your object is a TPDc or a TPDs (TPDcomm expected).")
  }
  if (!is.null(TPDs) & class(TPDs) != "TPDsp") {
    stop("Class mismatch: please specify if your object is a TPDc or a TPDs (TPDsp expected).")
  }

  results <- list()

  Calc_FRich <- function(x) {
    results_FR <- numeric()
    if (class(x) == "TPDcomm") {
      TPD <- x$TPDc$TPDc
      names_aux <- names(x$TPDc$TPDc)
      cell_volume <- x$data$cell_volume
    }
    if (class(x) == "TPDsp") {
      TPD <- x$TPDs
      names_aux <- names(x$TPDs)
      cell_volume <- x$data$cell_volume
    }
    for (i in 1:length(TPD)) {
      TPD_aux <- TPD[[i]]
      TPD_aux[TPD_aux > 0] <- cell_volume
      results_FR[i] <- sum(TPD_aux)
    }
    names(results_FR) <- names_aux
    return(results_FR)
  }

  if (!is.null(TPDc)) {
    results$communities <- list()
    results$communities$FRichness <- Calc_FRich(TPDc)
  }
  if (!is.null(TPDs)) {
    if (TPDs$data$type == "One population_One species" |
        TPDs$data$type == "One population_Multiple species") {
      results$species <- list()
      results$species$FRichness <- Calc_FRich(TPDs)
    } else {
      results$populations <- list()
      results$populations$FRichness <- Calc_FRich(TPDs)
    }
    if (TPDs$data$method == "mean") {
      message("WARNING: When TPDs are calculated using the TPDsMean function, Evenness and Divergence are meaningless!!")
    }
  }
  return(results)
}

imageTPD <- function(x, thresholdPlot = 0.99) {
  TPDList <- x$TPDc$TPDc
  imageTPD <- list()
  for (comm in 1:length(TPDList)) {
    percentile <- rep(NA, length(TPDList[[comm]]))
    TPDList[[comm]] <- cbind(index = 1:length(TPDList[[comm]]),
                             prob = TPDList[[comm]], percentile)
    orderTPD <- order(TPDList[[comm]][, "prob"], decreasing = TRUE)
    TPDList[[comm]] <- TPDList[[comm]][orderTPD, ]
    TPDList[[comm]][, "percentile"] <- cumsum(TPDList[[comm]][, "prob"])
    TPDList[[comm]] <- TPDList[[comm]][order(TPDList[[comm]][, "index"]), ]
    imageTPD[[comm]] <- TPDList[[comm]]
  }
  names(imageTPD) <- names(TPDList)

  trait1Edges <- unique(x$data$evaluation_grid[, 1])
  trait2Edges <- unique(x$data$evaluation_grid[, 2])
  imageMat <- array(
    NA,
    c(length(trait1Edges), length(trait2Edges), length(imageTPD)),
    dimnames = list(trait1Edges, trait2Edges, names(TPDList))
  )
  for (comm in 1:length(TPDList)) {
    percentileSpace <- x$data$evaluation_grid
    percentileSpace$percentile <- imageTPD[[comm]][, "percentile"]
    for (i in 1:length(trait2Edges)) {
      colAux <- subset(percentileSpace, percentileSpace[, 2] == trait2Edges[i])
      imageMat[, i, comm] <- colAux$percentile
    }
    imageMat[imageMat > thresholdPlot] <- NA
  }
  return(imageMat)
}

overlapF <- function(x, y) {
  if (!identical(x$eval.points, y$eval.points)) stop("Not identical eval points!")
  TPDx <- x$estimate
  if (class(x$eval.points) == "list") {
    x$eval.points <- expand.grid(x$eval.points)
  }
  cellEdge <- numeric()
  for (i in 1:ncol(x$eval.points)) {
    cellEdge[i] <- unique(x$eval.points[, i])[2] - unique(x$eval.points[, i])[1]
  }
  cellSizex <- prod(cellEdge)
  TPDx <- TPDx * cellSizex
  TPDx <- TPDx / sum(TPDx)

  TPDy <- y$estimate
  TPDy <- TPDy * cellSizex
  TPDy <- TPDy / sum(TPDy)

  OL <- sum(pmin(TPDx, TPDy))
  return(OL)
}

percentileTPD <- function(x) {
  TPDList <- x$TPDc$TPDc
  results <- list()
  for (comm in 1:length(TPDList)) {
    percentile <- rep(NA, nrow(TPDList[[comm]]))
    TPDList[[comm]] <- cbind(TPDList[[comm]], percentile)
    orderTPD <- order(TPDList[[comm]][, "notZeroProb"], decreasing = TRUE)
    TPDList[[comm]] <- TPDList[[comm]][orderTPD, ]
    TPDList[[comm]][, "percentile"] <- cumsum(TPDList[[comm]][, "notZeroProb"])
    TPDList[[comm]] <- TPDList[[comm]][order(TPDList[[comm]][, "notZeroIndex"]), ]
    results[[comm]] <- TPDList[[comm]]
  }
  names(results) <- names(TPDList)
  return(results)
}

quantileTPD <- function(x, thresholdPlot = 0.99) {
  TPDList <- x$TPDc$TPDc
  quantileTPD <- list()
  for (comm in 1:length(TPDList)) {
    percentile <- rep(NA, length(TPDList[[comm]]))
    TPDList[[comm]] <- cbind(index = 1:length(TPDList[[comm]]),
                             prob = TPDList[[comm]], percentile)
    orderTPD <- order(TPDList[[comm]][, "prob"], decreasing = TRUE)
    TPDList[[comm]] <- TPDList[[comm]][orderTPD, ]
    TPDList[[comm]][, "percentile"] <- cumsum(TPDList[[comm]][, "prob"])
    TPDList[[comm]] <- TPDList[[comm]][order(TPDList[[comm]][, "index"]), ]
    quantileTPD[[comm]] <- TPDList[[comm]]
  }
  names(quantileTPD) <- names(TPDList)

  trait1Edges <- unique(x$data$evaluation_grid[, 1])
  trait2Edges <- unique(x$data$evaluation_grid[, 2])
  imageMat <- array(
    NA,
    c(length(trait1Edges), length(trait2Edges), length(quantileTPD)),
    dimnames = list(trait1Edges, trait2Edges, names(TPDList))
  )
  for (comm in 1:length(TPDList)) {
    percentileSpace <- x$data$evaluation_grid
    percentileSpace$percentile <- quantileTPD[[comm]][, "percentile"]
    for (i in 1:length(trait2Edges)) {
      colAux <- subset(percentileSpace, percentileSpace[, 2] == trait2Edges[i])
      imageMat[, i, comm] <- 1 - colAux$percentile
    }
    imageMat[imageMat < (1 - thresholdPlot)] <- 0
  }
  return(imageMat)
}

MAE_TPD <- function(x, y) {
  diffXY <- abs(x - y)
  MAE <- mean(diffXY)
  return(MAE)
}
```

### Regex & snippets

```{r regex_and_snippets, eval=FALSE}
# Regex to select all symbols but '_'. useful for identify formulas in dplyr::
"[^[:alnum:][:space:]_]"

x <- iris
x <- x[, 1:2]
colnames(x) <- c("x1", "x2")

library(dplyr)

funfun <- function(data, expr = "log(x1 / x2) * 10^2") {
  expr <- parse(text = expr)
  data %>% mutate(x3 = eval(expr))
}

funfun(x)

## How to make a gower for big data table
library(cluster)       # daisy() and Gower distance
library(data.table)    # large data handling
library(parallel)      # parallel computing
library(bigstatsr)     # very large matrices

load(here::here("outputs", "test_phenofish_final.RData"))  # assumes object `test_phenofish_final`

data_morph <- test_phenofish_final %>%
  filter(trait_type == "morphological" & fresh == 1 & trait_coding == "continuous")

data_morph <- data_morph %>%
  dplyr::group_by(specimen_id, trait_name) %>%
  mutate(trait_value = as.numeric(trait_value))

data_morph <- data_morph %>%
  dplyr::group_by(phenofish_name, trait_name) %>%
  dplyr::summarise(trait_value = mean(trait_value), .groups = "drop")

species_traits <- reshape2::dcast(data_morph, phenofish_name ~ trait_name)
colnames(species_traits)[1] <- "species"
rownames(species_traits) <- species_traits[, 1]
traits <- species_traits[, -1]

block_size <- 10000
n <- nrow(traits)

gower_matrix <- matrix(NA, n, n)

compute_gower_block <- function(start_idx, end_idx, traits_data) {
  block_data <- traits_data[start_idx:end_idx, , drop = FALSE]
  gower_dist_block <- daisy(block_data, metric = "gower")
  return(as.matrix(gower_dist_block))
}

block_indices <- split(1:n, ceiling(1:n / block_size))

cl <- makeCluster(detectCores() - 1)
clusterExport(cl, varlist = c("traits", "daisy", "compute_gower_block"))

gower_blocks <- parLapply(cl, block_indices, function(indices) {
  start_idx <- indices[1]
  end_idx <- tail(indices, 1)
  compute_gower_block(start_idx, end_idx, traits)
})

for (i in 1:length(gower_blocks)) {
  idx_range <- block_indices[[i]]
  gower_matrix[idx_range, idx_range] <- gower_blocks[[i]]
}

stopCluster(cl)

print(gower_matrix[1:100, 1:100])
save(gower_matrix, file = "gower_matrix.RData")

pcoa_res <- ape::pcoa(as.dist(gower_matrix))
pca_res <- cmdscale(as.dist(gower_matrix), k = 2)
summary(pca_res)
```
